---
title: "Figure 1. Single-Cell and Tissue Phenotype prediction (SCTP). "
author: "Wencan Zhu"
output: html_document
date: "2025-06-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data loading

```{r function}
library(ggplot2)
library(Seurat)
library(RColorBrewer)
source("~/Documents/GitHub/SCTP_iMETA/Fig1/Eval_func.R")
```

## TSNE representation of the single cell database used for prediction. Cells are colored by global type and subtype annotations respectively. 
```{r SC data cell types}
load("/Users/w435u/Documents/ST_SC/OUTPUT2_test/Seurat_SC_pred.RData")
col=brewer.pal(n = 11, name = "RdBu")

my_cols <- c("Epithelial cells"='#A4DFF2',
             "Mast cells"='#5C88DAFF',
             "Myeloids"='#00B5E2FF',
             "Stromal cells"='#7C878EFF',
             "T cells" ='#ff9a00',
             "B cells"='#CC0C00FF')

DimPlot(pred_sc_seurat, group.by = "Global_Cluster", reduction = "tsne", cols = my_cols)

out_name = "~/Documents/GitHub/SCTP_iMETA/Fig1/SC_DimPlot_Global_Cluster.pdf"
pdf(out_name)
DimPlot(pred_sc_seurat, group.by = "Global_Cluster", reduction = "tsne", cols = my_cols)
dev.off()
```

## Prediction of cell phenotype (malignant vs. non-malignant) based on different spatial transcriptomic datasets by SCTP.
```{r SC data prediction}
DimPlot(pred_sc_seurat, reduction = "tsne", group.by = c('sign_sc','sign_sc2','sign_sc3','sign_sc4' ), cols =  c(col[10], col[2]),  pt.size=0.00001)+ggtitle("")
ggsave("~/Documents/GitHub/SCTP_iMETA/Fig1/SC_DimPlot_pred_all.pdf")


```
## Nonimmune cell malignancy prediction based on SCTP-CRC. TSNE presentation of cells, with cells colored by the predicted phenotype from SCTP-CRC


```{r SC data prediction CAO}
load("/Users/w435u/Documents/ST_SC/DATA_STSC_CAO/Seurat_nonimmune.RData")
load("/Users/w435u/Documents/ST_SC/OUTPUT_CAO_new/Meta_pred_sc_CAO_degree100.RData")
sc_nonimmune_pred <- AddMetaData(sc_nonimmune_seurat, metadata = metadata_sc100)
name_meta <- paste0("sign_", "C1", "_100")
pdf(paste0("~/Documents/GitHub/SCTP_iMETA/Fig1/TSNE_pred_SC", "C1", "_100.pdf"))
gg <- DimPlot(sc_nonimmune_pred, reduction = "tsne", group.by = name_meta, cols =  c("#053061", '#EB455F'), 
          order=c("negative","positive"), pt.size=0.0000001)+NoLegend() +ggtitle("")
print(gg)
dev.off()
  
print(gg)
```

## True labels and SCTP-HCC predictions on independent HCC dataset. 
```{r SCTP prediction}
label_colors <- c("tumor" = "#E075A0", "normal" ="#77aecb")

fig_all <- label_all <- list()
for (i in 1:4) {
  load(paste0("/Users/w435u/Documents/ST_SC/OUTPUT_HCC/spatial/ST-liver",i, "_seurat.RData"))
  fig_all[[i]] <- SpatialFeaturePlot(pred_st_dataset, "malignancy",  pt.size.factor = 2.5)+
 scale_fill_gradientn(colours = rev(brewer.pal(n = 11, name = "RdBu")))

  label_all[[i]] <- SpatialDimPlot(pred_st_dataset, "label",  cols = label_colors,, pt.size.factor = 2.5 )
  print(fig_all[[i]])
  print(label_all[[i]])
}

```


## Benchmarking prediction performance of SCTP and other three existing methods on independent HCC dataset, including accuracy, F1 score, AUC and Brier score. 

```{r result comparison HCC, echo=TRUE}
eval_all =  matrix(NA, 4, 8)
eval_cf = eval_copykat = eval_meti = matrix(NA, 4, 8)

plot_methods <- c( "malignancy", "pred_meti", "pred_ck", "pred_cf")
for (i in (1:4)) {
  out_cf <- read.csv(paste0("~/Documents/GitHub/SCTP_iMETA/Fig1/out_HCC_liver",i,".csv"))
   pred_cf <- out_cf$predict
  copykat_prediction <- read.delim(paste0("~/Documents/GitHub/SCTP_iMETA/Fig1/ST-liver",i, "_copykat_prediction.txt"))
   rownames(copykat_prediction) <- copykat_prediction$cell.names
  #copykat_prediction$pred_cnv <- ifelse(substr(copykat_prediction$copykat.pred,1,2)=="c2", 1, 0)
  copykat_prediction$pred_cnv <- ifelse(substr(copykat_prediction$copykat.pred,1,2)=="an", 1, 0)
  #test <- merge(copykat_prediction, pred_st_dataset@meta.data, by=0, all=TRUE)
  
  METI_label <- read.csv(paste0("~/Documents/GitHub/SCTP_iMETA/Fig1/METI_label_liver",i, ".csv"), row.names=1)
  test <- merge(copykat_prediction, METI_label, by=0, all=TRUE)
  pred_meti <- ifelse(METI_label$label=="Tumor", 1, 0)
  SCTP_pred_liver1 <- read.csv(paste0("~/Documents/GitHub/SCTP_iMETA/Fig1/SCTP_pred_liver",i, ".csv"), row.names=1)
  test$pred_sctp <-  SCTP_pred_liver1
  test$pred_meti <- ifelse(METI_label$label=="Tumor", 1, 0)
  test$pred_cf <- pred_cf
  true_labels <- read.csv(paste0("~/Documents/GitHub/SCTP_iMETA/Fig1/ST_liver", i, "_true_labels.csv"), row.names=1)
  
  label <- ifelse(true_labels=="tumor", 1, 0 )
  
  eval_copykat[i, ] <- func_Eval(test$pred_cnv, label)
  eval_meti[i, ] <- func_Eval(test$pred_meti, label)
  eval_all[i, ] <- func_Eval(test$pred_sctp$x, resp_vec = label)
  eval_cf[i, ] <- func_Eval(test$pred_cf, label)
  #eval_all_old[i, ] <- func_Eval(pred_old, label)
}

my_blue <- "#095786"
my_red <- "#ae0000"
my_colors <- c("#E57470","#FBD37E", "#7A7DEE", "#A6DBFB")
df_plot <- data.frame(rbind(eval_all, eval_cf,eval_meti, eval_copykat))
colnames(df_plot) <- c("AUC", "AUCPR", "Brier", "F1","ACC", "LogLoss", "Sensitivity")
df_plot$method <- c(rep("SCTP",4), rep("CancerFinder", 4),rep("METI", 4), rep("copykat", 4))
df_plot$method <- factor(df_plot$method, levels = c("SCTP", "METI", "copykat", "CancerFinder"))
df_plot$data <- rep(c("HCC1", "HCC2","HCC3","HCC4"),4)

g_auc <- ggplot(df_plot, aes(data, AUC, group = method, fill=method))+
  geom_col( position = 'dodge',width = 0.8, alpha=1)+theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.border = element_blank())+
  scale_y_continuous(breaks=c(seq(from=0,to=1,by=0.2)))+
  scale_fill_manual(values = my_colors)

g_f1 <- ggplot(df_plot, aes(data, F1, group = method, fill=method))+
  geom_col( position = 'dodge',width = 0.8, alpha=1)+theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.border = element_blank())+
  scale_y_continuous(breaks=c(seq(from=0,to=1,by=0.2)))+
  scale_fill_manual(values = my_colors)


g_acc <- ggplot(df_plot, aes(data, ACC, group = method, fill=method))+
  geom_col( position = 'dodge',width = 0.8, alpha=1)+theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.border = element_blank())+
  scale_y_continuous(breaks=c(seq(from=0,to=1,by=0.2)))+
  scale_fill_manual(values = my_colors)

g_sens <- ggplot(df_plot, aes(data, Sensitivity, group = method, fill=method))+
  geom_col( position = 'dodge',width = 0.8, alpha=1)+theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.border = element_blank())+
  scale_y_continuous(breaks=c(seq(from=0,to=1,by=0.2)))+
  scale_fill_manual(values =my_colors)

g_Brier <- ggplot(df_plot, aes(data, Brier, group = method, fill=method))+
  geom_col( position = 'dodge',width = 0.8, alpha=1)+theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.border = element_blank())+
  scale_y_continuous(breaks=c(seq(from=0,to=1,by=0.2)))+
  scale_fill_manual(values = my_colors)

library(ggpubr)

g <- ggarrange(
  g_acc,g_auc,g_f1, g_Brier , ncol=4, 
  common.legend = TRUE, legend = "bottom"
)

pdf(file="~/Documents/GitHub/SCTP_iMETA/Fig1/Eval_all_stat_new.pdf", width=12, height=4)
print(g)
dev.off()

print(g)

```

```{r }
save(df_plot, file= "~/Documents/GitHub/SCTP_iMETA/Fig1/DF_plot_eval.RData")
```



Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
